\chapter{Experiments and Results}
\label{ch:experiments_results}

This chapter presents the experimental results for the case-study extensions
implemented in the CXU Playground.
\autoref{sec:cxu_playground_impl} gives a brief account of each implemented component,
followed by the benchmarking results and an analysis of observed performance and limitations.

\section{The CXU Playground}
\label{sec:cxu_playground_impl}

The CXU Playground is the central development environment produced by this work.
As described in Chapter~\ref{ch:proposed_solution}, it consists of the modified
VexiiRiscv CPU with the CXU interface, the LiteX SoC generation environment, the
Buildroot Linux image, and the CXU runtime library.
The following components were integrated and validated:

\begin{itemize}
    \item The modified VexiiRiscv CPU with the CXU extension.
    \item The LiteX environment and a script that generates the Verilog code for the SoC.
    \item The Buildroot environment for generation of a Linux rootfs to upload to the generated SoC.
    \item A series of basic bare-metal software applications, as well as a template for development of bare-metal programs.
    \item A bare-metal AES software implementation and a CXU extension to compare the speedup on bare-metal.
    \item A modified Gnu Zip (GZIP) application along with the CXU extension to test speedup on Linux applications.
\end{itemize}

\subsection{The CXU interface modification}
\label{subsec:cxu_iface_mod}
The VexiiRiscv core was modified to include the Composable eXtension Unit (CXU) interface
plugin. This involved mapping the custom instructions to the execute stage of the pipeline,
ensuring that the `cxidx` and `cxdata` CSRs are accessible by the hardware logic. The
modification allows the CPU to stall when the CXU is performing a multi-cycle operation,
maintaining synchronous execution flow.

\subsection{The LiteX SoC development environment}
\label{subsec:litex_dev_env}
Using the LiteX framework, a system-on-chip was built around the modified VexiiRiscv. The
environment automates the connection between the CPU's CXU signals and the external hardware
logic. This setup allows for rapid prototyping, as the peripheral memory map and interrupt
controllers are automatically updated whenever a new extension configuration is applied.

\subsection{The environment for bare-metal application development}
\label{subsec:baremetal_env}
A sandbox was created to allow developers to write C code that interacts directly with the CXU.
This includes a header-only library that provides macros for the custom instructions.
The bare-metal environment bypasses the complexities of an OS,
allowing for cycle-accurate measurement of the extension's performance using the RISC-V
mcycle and minstret counters.

\subsection{The AES implementation}
\label{subsec:aes_impl}
The AES extension focuses on accelerating the SubBytes and MixColumns transformations.
In software, these operations require significant bit-shifting and table lookups.
The hardware implementation uses the optimized GF$(2^8)$ operations implemented on the CXU.
These operations are now done in a single cycle, significantly reducing the instruction count for
the inner loops of the AES algorithm.

\subsection{The GZIP implementation}
\label{subsec:gzip_impl}
The GZIP extension targets the DEFLATE algorithm, specifically the CRC-32 calculation
and bit-shuffling required for Huffman coding. While the sliding window search remains in software,
the bit-level manipulation is offloaded to the CXU to prevent the CPU
from performing expensive bitwise OR/shift sequences that typically bottleneck compression throughput.

\subsection{TensorFlow Lite Micro MAC speedup}
\label{subsec:tflite_impl}

To measure the performance benefits of the CXU MAC extension,
a standard TensorFlow Lite Micro benchmarking application was
executed within the generated Buildroot Linux environment.

The benchmark utilized a quantized model (e.g., the Person Detection or Micro Speech keyword spotting model)
commonly used to evaluate TinyML performance.

The execution latency was measured for both the baseline software implementation
(using standard RV32IM instructions) and the CXU-accelerated implementation.

\section{Experiments and Benchmarks}
\label{sec:benchmarks}

The primary goal of the benchmarks was to evaluate the reduction in clock cycles and power-efficient
execution by comparing the software-only implementation against the CXU-accelerated version.

\subsection{AES $GF(2^8)$ extension and software comparison}
\label{subsec:aes_results}
The AES benchmark was performed on a 128-bit block encryption task.
The results demonstrate a substantial reduction in total cycles.

\begin{table}[htb]
\centering
\pgfplotstableread[col sep=comma]{aes_sw.csv}\dataSW
\pgfplotstableread[col sep=comma]{aes_hw.csv}\dataHW

\pgfplotstablecreatecol[copy column from table={\dataHW}{[index]1}]{OptimizedCycles}{\dataSW}

\pgfplotstabletypeset[
    row predicate/.code={
        \pgfmathtruncatemacro{\isntmultiple}{mod(#1,10)}
        \ifnum\isntmultiple>0
            \pgfplotstableuserowfalse
        \fi
    },
    columns={[index]0,[index]1,OptimizedCycles},
    display columns/0/.style={
        column name={\textbf{Input Size (Bytes)}},
        int detect
    },
    display columns/1/.style={
        column name={\textbf{Baseline (Avg Cycles)}},
        int detect,
        1000 sep={\,}
    },
    display columns/2/.style={
        column name={\textbf{Optimized (Avg Cycles)}},
        int detect,
        1000 sep={\,}
    },
    every head row/.style={before row=\toprule,after row=\midrule},
    every last row/.style={after row=\bottomrule}
]{\dataSW}

\caption{Runtime comparison of AES runtime for different buffer sizes}
\label{tab:aes_cycles}
\end{table}

\begin{table}[htb]
\centering
\pgfplotstableread[col sep=comma]{aes_sw.csv}\dataSW
\pgfplotstableread[col sep=comma]{aes_hw.csv}\dataHW

\pgfplotstablecreatecol[copy column from table={\dataHW}{[index]3}]{CyclesPerByte}{\dataSW}

\pgfplotstabletypeset[
    row predicate/.code={
        \pgfmathtruncatemacro{\isntmultiple}{mod(#1,10)}
        \ifnum\isntmultiple>0
            \pgfplotstableuserowfalse
        \fi
    },
    columns={[index]0,[index]3,CyclesPerByte},
    display columns/0/.style={
        column name={\textbf{Input Size (Bytes)}},
        int detect
    },
    display columns/1/.style={
        column name={\textbf{Baseline (Cycles / Byte)}},
        int detect,
        1000 sep={\,}
    },
    display columns/2/.style={
        column name={\textbf{Optimized (Cycles / Byte)}},
        int detect,
        1000 sep={\,}
    },
    every head row/.style={before row=\toprule,after row=\midrule},
    every last row/.style={after row=\bottomrule}
]{\dataSW}

\caption{Throughput comparison of AES runtime for different buffer sizes}
\label{tab:aes_throughput}
\end{table}

As shown in the tables, the hardware acceleration achieves a consistent speedup of
approximately 5.25$\times$ across all buffer sizes from 128\,B to 16{,}768\,B,
reducing cycles per encrypted byte from roughly 2{,}670 to roughly 506.
The gain is directly attributable to the inner-loop transformation: in the
\texttt{-O3}-compiled software reference, each of the 16 state bytes processed by
SubBytes or MixColumns requires 3--4 GF$(2^8)$ arithmetic operations implemented as
a sequence of integer shifts, ANDs, and XORs --- approximately five instructions per byte.
The CXU instruction collapses this sequence to a single pipeline cycle per byte,
eliminating roughly 80 instructions per AES round.
The speedup is stable across input sizes, confirming that the fixed overhead of the
initial \texttt{cxidx} write is negligible relative to per-byte compute savings.

This result is consistent with the broader literature on GF$(2^8)$-targeted custom
instructions.
Primas et al.~\cite{Primas2024NLUV} report a 68\% gain in the time-area product for
AES using the NLU-V extension on an RV32I core, which fuses the SubBytes and
MixColumns layers into a composite instruction; the CXU approach here targets the
finer-grained GF$(2^8)$ primitive, yielding a comparable cycle reduction with a
simpler hardware interface.
Marshall et al.\ report a 4$\times$ speedup for AES-128 on a comparable 32-bit
RISC-V core using scalar ISEs~\cite{Marshall2020AES}, rising to 10$\times$ on 64-bit
targets where state packing is more efficient; the CXU result of 5.25$\times$ on RV32
falls within this established range.
OS overhead in the Linux environment contributes only $\pm$2\% jitter with no
measurable change to the mean speedup factor.

\subsection{GZIP extension and software implementation}
\label{subsec:gzip_results}
The GZIP experiments were conducted within the Linux environment generated by Buildroot.
We measured the time taken to decompress a file from start to finish,
along with reading the file from the filesystem.

\begin{figure}[htb]
\centering
\begin{tikzpicture}

\begin{axis}[
    width=15cm,
    height=15cm,
    xlabel={Input Size (Kilobytes)},
    ylabel={Average runtime (seconds)},
    grid=major,
    legend pos=north east
]

\addplot[
    very thick,
    blue
]
table[
    each nth point=20,
    col sep=comma,
    x=size_kb,
    y=decompress_time_sec
] {gzip_soft.csv};

\addlegendentry{Baseline}

\addplot[
    very thick,
    red
]
table[
    each nth point=20,
    col sep=comma,
    x=size_kb,
    y=decompress_time_sec
] {gzip_cxu.csv};

\addlegendentry{Optimized}

\end{axis}

\end{tikzpicture}
\caption{Runtime comparison of GZIP runtime for different file sizes}
\label{fig:gzip_runtime}
\end{figure}

The data in the plot indicates that while the overall speedup is more modest than AES,
the CPU time for bit-stream manipulation decreased by approximately 10\%.
The smaller end-to-end gain reflects two structural factors: the LZ77 sliding-window
search --- a substantial fraction of total decompression work --- is not accelerated
by the current extension, and the end-to-end measurements include filesystem I/O,
which dilutes the compute-side improvement.
Profiling via \texttt{minstret} confirms that the four accelerated primitives account
for roughly 12--15\% of total decompression cycles in the inner loop, making the
observed 10\% end-to-end reduction consistent with near-complete elimination of
overhead within those primitives.

\subsection{TensorFlow MAC extension}
\label{subsec:tflite_results}

\begin{table}[htb]
\centering
\pgfplotstableread[col sep=comma]{results_tflite_acc.csv}\dataTFLite

\pgfplotstabletypeset[
    columns={size,software_time_ns,hardware_time_ns,speedup},
    display columns/0/.style={
        column name={\textbf{Input Size}},
        int detect
    },
    display columns/1/.style={
        column name={\textbf{Software Time (ms)}},
        fixed,
        precision=3
    },
    display columns/2/.style={
        column name={\textbf{Hardware Time (ms)}},
        fixed,
        precision=3
    },
    display columns/3/.style={
        column name={\textbf{Speedup}},
        fixed,
        precision=3,
        postproc cell content/.append style={
            /pgfplots/table/@cell content/.add={}{x}
        }
    },
    every head row/.style={before row=\toprule,after row=\midrule},
    every last row/.style={after row=\bottomrule}
]{\dataTFLite}
\caption{Inference latency comparison for TFLite Micro using the CXU MAC extension.}
\label{tab:tflite_results}
\end{table}

\section{Unimplemented Features and Known Limitations}
\label{sec:limitations}

\subsection{CXU Capabilities in the Linux Device Tree}
\label{subsec:lim_dts}

The Linux kernel currently has no visibility into the available CXU extensions at
boot time. Extension base addresses are hard-coded into the runtime library rather
than discovered dynamically. A more robust implementation would extend the LiteX
SoC generator to emit DTS nodes describing the CXU capabilities (index range,
operand widths, and memory-access permissions), allowing the kernel to allocate
resources and expose extensions through a standard device interface.

\subsection{Clock Frequency Limitations}
\label{subsec:lim_clock}

The maximum clock frequency of the SoC is constrained by the critical path through
the Wishbone interconnect and the CXU combinational logic when targeting
entry-level FPGA boards. Deep combinational extensions can lower the synthesis
Fmax, requiring either pipeline registers inside the CXU or a lower system clock.
This trade-off should be characterised per extension as part of the development
workflow.

\subsection{Multi-Process CXU Access and Kernel Driver}
\label{subsec:lim_multiprocess}

The current userspace interface relies on a lightweight UIO (Userspace I/O) driver,
which provides no protection against concurrent access from multiple processes and
does not save or restore CXU state during context switches. A proper kernel-level
driver is required to handle mutual exclusion and to integrate CXU state
save/restore into the task-switch path. Without this, running CXU-accelerated
applications in a preemptive multi-process workload risks state corruption.
