\chapter{Introduction}
\label{ch:introduction}

\section{Motivation}
\label{sec:motivation}

RISC-V \cite{RISC-VInternational2025} is a modern and extensible instruction set architecture often used in systems-on-chip,
especially, FPGA-based ones \cite{LiteXGitHub}, such that it is both architecturally simple and energy-efficient.
RISC-V's instruction set architecture is designed with custom extensions in mind,
so designing custom extensions would be a straightforward process.
It is, however, a difficult task to integrate a custom extension into the RISC-V specification,
and even more so to implement an extension for the multitude of RISC-V processor implementations.
With such issues, and the fact that most FPGA-based SOCs offer plenty
of space to also implement an extension,
in mind, Jan Gray et al. designed the custom composable extensions (CFU) \cite{Gray2022CFU} extension.
It allows for an implementation of a specific extension to be used across multiple RISC-V CPUs,
so as to use the single extension implementation across a multitude of different RISC-V CPUs.
CFU allows for a single extension to be connected to a hardware thread.
It also allows passing a ''function index'' to an extension, such that up to 255 different functions may be implemented.

There is also a playground for desgining the CFU extensions on a ready-made SOC called
the CFU playground. It allows the users to implement a
CFU and write a main script that runs and invokes the implemented extension.
CFU is a useful interface for implementing a single extension,
but when combining extensions, especially those,
that may have multiple different functions and utilize the function id,
managing them becomes complex and introduces significant software overhead.
Because of that, Gray et al. also designed a newer version of composable
custom extensions - the CXU \cite{Gray2025CXU}.
This implementation allows for switching between multiple custom extensions,
but, as of now, there is not yet an implementation of it
on any of the RISC-V implementations, nor an environment for developing custom extensions.

\section{Contributions}
\label{sec:contributions}

This thesis makes the following concrete contributions:

\begin{itemize}
    \item \textbf{First CXU implementation:} A SpinalHDL plugin for the
    VexiiRiscv processor that realises the CXU interface, including opcode
    interception at the execute stage, pipeline stall management for multi-cycle
    operations, and \texttt{cxidx}/\texttt{cxdata} CSR support.

    \item \textbf{CXU Playground:} A unified, open-source sandbox built on LiteX
    that automates SoC generation, Buildroot Linux integration, cross-compilation,
    and CXU hardware templating, enabling developers to prototype and benchmark
    custom extensions under realistic, OS-level conditions.

    \item \textbf{C-level CXU runtime library:} A header-only library that wraps
    the low-level assembly invocation, manages extension multiplexing via the
    \texttt{cxidx} register, and provides a portable C API for CXU-accelerated
    routines.

    \item \textbf{Three case-study extensions:} An AES GF$(2^8)$ accelerator, a
    GZIP/DEFLATE bit-manipulation accelerator, and a multiply-accumulate (MAC)
    extension for TensorFlow Lite Micro inference, each accompanied by a paired
    software reference implementation and benchmarking suite.
\end{itemize}

Benchmarking results demonstrate approximately a 5.25$\times$  reduction in cycles
per byte for AES encryption and a roughly 10\% reduction in CPU time for GZIP
bit-stream manipulation, confirming that even a small number of tightly coupled
custom instructions can yield substantial gains for arithmetic-intensive workloads (see \autoref{ch:experiments_results}).

\section{Structure Of The Thesis}
\label{sec:structure}

\begin{itemize}
    \item \autoref{ch:related_work} surveys existing custom-instruction interfaces (CFU,
    CXU, Core-V-XIF), FPGA-based SoC frameworks, related sandboxes, and prior
    work applying custom ISA extensions to cryptography, compression, and machine
    learning.

\item \autoref{ch:theoretical_background} provides the theoretical background covering the
    RISC-V ISA, the VexiiRiscv pipeline, the LiteX SoC ecosystem, the CXU
    coupling mechanism, Linux boot flow on LiteX, and the algorithmic foundations
    of the three case studies.

\item \autoref{ch:proposed_solution} describes the design of the CXU Playground: the
    system architecture, the CXU plugin integration in VexiiRiscv, the Buildroot
    environment, the runtime library, and the benchmarking methodology.

\item \autoref{ch:experiments_results} presents the implementation details and experimental
    results for the AES, GZIP, and TFLite Micro extensions, and discusses
    unimplemented features and known limitations.

\item \autoref{ch:conclusions} summarises the contributions, reflects on the
    limitations of the current implementation, and outlines directions for future
    work.
\end{itemize}
