\chapter{Proposed Solution}
\label{ch:proposed_solution}

\section{Problem Statement}
\label{sec:problem_statement}

% O.F. Трохи є повторів з попереднім, але,  принципі, хай буде...
The RISC-V ISA is designed with the explicit intent of supporting custom extensions,
allowing developers to optimise processors for specific workloads.
In practice, however, the process of designing, integrating, and evaluating custom
instructions remains highly complex and implementation-specific.
The CFU specification~\cite{Gray2022CFU} addressed reusability across cores, but its
single-extension-per-thread model and assembly-only invocation made it impractical for
more demanding use cases.
The CXU specification~\cite{Gray2025CXU} was proposed to resolve these issues, offering
support for multiple simultaneous extensions, shared state, and C-level invocation via
a standardized ABI.
As established in Chapter~\ref{ch:related_work}, however, to the best of our knowledge, no hardware implementation
of the CXU interface existed prior to this work, and no sandbox environment supported
its features.

The main focus of this work is therefore a concrete implementation of the CXU interface
along with a playground for prototyping and development of custom CXU extensions.
A dedicated sandbox environment reduces the development time
required to test specialized hardware accelerators,
such as those needed for cryptography and data compression.

Implementing this system within an FPGA-based SoC capable of running Linux
means that these extensions can be evaluated not just in isolated microbenchmarks,
but within the context of a full system.

Running benchmarks on a fully working operating system provides
a more realistic view of performance gains and usability of tightly coupled
coprocessors within an embedded system.

\section{System Architecture and Playground Design}
\label{sec:system_arch}

% O.F. Знову повтори... Можливо, щось з цього мало б перекочувати у попередній розділ -- теорії
The CXU Playground is built on the LiteX framework~\cite{LiteXGitHub}
to focus effort to focus on the CXU integration rather than peripheral assembly.
LiteX enables automated generation of the FPGA-based SoC by connecting the CPU core
to essential peripherals --- including UART and Ethernet --- and to external memory via
LiteDRAM.
Communication between these modules is handled by the Wishbone bus, which provides
memory-mapped I/O access to all connected components, as described in
Section~\ref{subsec:interconnect}.

The core processor of the SoC is VexiiRiscv~\cite{VexiiRiscVGithub}, configured as an
RV32IMAFC target and implemented in SpinalHDL. Integer operations and atomics are used for linux,
and compressed instructions / floats are used to simplify compiling TensorFlow ~\cite{David2021TFLiteMicro}.
% O.F. Тут на пару букв більше, ніж в попередньому розділі -- може поясніть
VexiiRiscv's plugin-based architecture allows the CXU extension to be integrated as an
independent execute-stage plugin without modifying the core pipeline, as discussed in
Section~\ref{subsec:vexii}.

The entire playground is available as a single open-source repository~\cite{CXUPlayground}.
The repository is structured to separate hardware generation, software toolchains, and
user-defined custom extensions, and provides scripts for automating the LiteX SoC build
process, templates for writing SpinalHDL extensions, and a structured environment for
compiling and running software that invokes custom instructions.

\section{CXU Integration in VexiiRiscv}
\label{sec:cxu_integration}

To provide low-latency instruction-level acceleration, the CXU mechanism was integrated
as a tightly coupled accelerator directly within the CPU pipeline, operating as a user
plugin during the execute stage of VexiiRiscv.
Because VexiiRiscv is written in SpinalHDL, the integration was implemented as a
dedicated pipeline plugin that intercepts specific instruction streams.
The data flow through this plugin is as follows: opcode patterns reserved for custom
instructions are mapped to the CXU during the decode stage; the CXU command then
receives source operands directly from CPU registers \textbf{rs1} and \textbf{rs2};
the custom logic processes these inputs based on its internal state and returns the
output to the destination register \textbf{rd}; and software manages and queries the
CXU's shared state via the dedicated \textbf{cxidx} (Extension Index) and
\textbf{cxdata} Control/Status Registers (CSRs).
For multi-cycle operations, the plugin freezes the CPU pipeline,
managing pipeline stalls and structural hazards to ensure correct execution ordering
and data integrity.
% O.F. Мені тут знань бракує -- саме halt signal? Не stall абощо? 
% P. M. Тут навіть не halt, а буквально заморожує конвеєр, це просто назвали також halt в SpinalHDL, вибачте >~<

Figure~\ref{fig:cxu_pipeline} illustrates how the CXU plugin sits below the Execute
stage of the pipeline, receiving operands from Decode and returning results that are
written back at the Execute/Memory boundary.

\begin{figure}[htb]
    \centering
    \begin{tikzpicture}[
        node distance=0.8cm and 0.5cm,
        stage/.style={draw, rectangle, minimum width=2cm, minimum height=1.2cm, fill=blue!10, rounded corners, font=\bfseries},
        cxu/.style={draw, align=center, rectangle, minimum width=3cm, minimum height=1.5cm, fill=green!10, rounded corners, font=\bfseries},
        arrow/.style={-Latex, thick},
        signal/.style={-Latex, dashed, thick, red}
    ]
    \node[stage] (fetch) {Fetch};
    \node[stage, right=of fetch] (decode) {Decode};
    \node[stage, right=of decode] (execute) {Execute};
    \node[stage, right=of execute] (memory) {Memory};
    \node[stage, right=of memory] (writeback) {Writeback};

    \node[cxu, below=1.5cm of execute] (cxu_unit) {CXU Plugin \\ \small (Custom Accelerator)};

    \draw[arrow] (fetch) -- (decode);
    \draw[arrow] (decode) -- (execute);
    \draw[arrow] (execute) -- (memory);
    \draw[arrow] (memory) -- (writeback);

    \draw[arrow] (decode.south) -- ++(0,-0.25) -| node[pos=0.75, left, font=\small] {Opcode Map} ([xshift=-1cm]cxu_unit.north);
    \draw[arrow] ([xshift=-0.4cm]execute.south) -- node[pos=0.5, above, font=\small] {rs1, rs2} ([xshift=-0.4cm]cxu_unit.north);
    \draw[arrow] ([xshift=0.4cm]cxu_unit.north) -- node[right, font=\small] {rd} ([xshift=0.4cm]execute.south);

    \draw[signal] (cxu_unit.east) -| node[pos=0.75, right, font=\small, text=black] {Halt / Pipeline Stall} (memory.south);

    \end{tikzpicture}
    \caption{Integration of the CXU as a tightly coupled plugin within the VexiiRiscv pipeline.}
    \label{fig:cxu_pipeline}
\end{figure}

\section{Sandbox Development Environment and Buildroot Integration}
\label{sec:sandbox_env}

The core contribution of this work is the unified sandbox that enables developers to
more easily design, test, and benchmark CXU hardware.
Building software for the LiteX-based system requires a cross-compilation toolchain
targeting the RV32IMA instruction set and ILP32 ABI, as described in
Section~\ref{subsec:cross_compile}.

To evaluate custom extensions under realistic conditions, the sandbox supports running
a full Linux operating system.
This is achieved through Buildroot integration, which cross-compiles the Linux kernel,
generates the root filesystem, and builds the essential user-space utilities required
for benchmarking.
Linux compatibility requires configuring the VexiiRiscv core with an MMU, supervisor
privilege mode, and atomic instruction support, as established in
Section~\ref{subsec:hw_requirements}.
During the build process, LiteX automatically generates the Device Tree Source (DTS),
which is compiled into a DTB and passed to the Linux kernel by the OpenSBI bootloader,
dynamically mapping the memory layout, hardware peripherals, and --- in future work ---
CXU capabilities to the operating system.
Figure~\ref{fig:cxu_stack} shows the resulting full-system software stack.

% O.F. Через невирівняність картинка, звичайно, страшненька вийшла... Але немає ради, хай вже.
\begin{figure}[htb]
    \centering
    \begin{tikzpicture}[
        layer/.style={draw, rectangle, minimum width=8.5cm, minimum height=1.2cm, font=\bfseries},
        hw/.style={fill=gray!20},
        fw/.style={fill=purple!20},
        os/.style={fill=orange!20},
        rt/.style={fill=cyan!10},
        usr/.style={fill=cyan!30},
        box/.style={rectangle, draw, rounded corners=3pt,
            minimum width=10cm, minimum height=0.75cm, align=center,
            text centered, font=\small}
    ]
    \node[box, hw] (hardware) {Hardware: LiteX SoC + VexiiRiscv + CXU Extensions};
    \node[box, fw, above=0cm of hardware] (firmware) {Firmware: OpenSBI \& Device Tree Blob (DTB)};
    \node[box, os, above=0cm of firmware] (linux) {Operating System: Linux Kernel (MMU enabled)};
    \node[box, rt, above=0cm of linux] (runtime) {CXU Runtime ABI \& CXU Library};
    \node[box, usr, above=0cm of runtime] (apps) {User Space Applications (AES, GZIP Benchmarks)};

    \node[left=0.8cm of hardware, align=right, font=\itshape] {Kernel Mode};
    \node[left=0.8cm of runtime, align=right, font=\itshape] {User Mode};

    \end{tikzpicture}
    \caption{The CXU Sandbox full-system stack.}
    \label{fig:cxu_stack}
\end{figure}

\section{The CXU Runtime}
\label{sec:cxu_runtime}

Bringing CXU extensions to an operating system environment requires a software
interface layer that handles extension selection and, in principle, state preservation
across context switches.
When multiple applications running on Linux utilize the CXU, the state of the custom
hardware unit must be preserved so that a preempted process can resume without
corruption.

The CXU runtime addresses this through a standardized Application Binary Interface
(ABI) and a header-only C-level library.
The library wraps the low-level assembly instructions into safe, callable C functions
and manages the \textbf{cxidx} register to multiplex access to different hardware
extensions among software threads.
It is important to note that, while the CXU specification defines the save/restore
semantics required for OS context switching, the present implementation has not yet
integrated this into the Linux kernel's task-switch path.
As a consequence, the current system is safe only for single-process CXU workloads
under Linux; the multi-process correctness gap is discussed as a limitation in
Section~\ref{subsec:lim_multiprocess}.

\section{The CXU Playground}
\label{sec:cxu_playground_impl}

The CXU Playground is the central development environment produced by this work.
As described in Chapter~\ref{ch:proposed_solution}, it consists of the modified
VexiiRiscv CPU with the CXU interface, the LiteX SoC generation environment, the
Buildroot Linux image, and the CXU runtime library.
The following components were integrated and validated:

\begin{itemize}
    \item The modified VexiiRiscv CPU with the CXU extension.
    \item The LiteX environment and a script that generates the Verilog code for the SoC.
    \item The Buildroot environment for the generation of a Linux rootfs to upload to the generated SoC.
    \item A series of basic bare-metal software applications, as well as a template for development of bare-metal programs.
    \item A bare-metal AES software implementation and a CXU extension to compare the speedup on bare metal.
    \item A modified Gnu Zip (GZIP) application along with the CXU extension to test speedup on Linux applications.
\end{itemize}

\subsection{The CXU interface modification}
\label{subsec:cxu_iface_mod}
The VexiiRiscv core was modified to include the Composable eXtension Unit (CXU) interface
plugin. This involved mapping the custom instructions to the execute stage of the pipeline,
ensuring that the `cxidx` and `cxdata` CSRs are accessible by the hardware logic. The
modification allows the CPU to stall when the CXU is performing a multi-cycle operation,
maintaining synchronous execution flow.

\subsection{The LiteX SoC development environment}
\label{subsec:litex_dev_env}
Using the LiteX framework, a system-on-chip was built around the modified VexiiRiscv. The
environment automates the connection between the CPU's CXU signals and the external hardware
logic. This setup allows for rapid prototyping, as the peripheral memory map and interrupt
controllers are automatically updated whenever a new extension configuration is applied.

\subsection{The environment for bare-metal application development}
\label{subsec:baremetal_env}
A sandbox was created to allow developers to write C code that interacts directly with the CXU.
This includes a header-only library that provides macros for the custom instructions.
The bare-metal environment bypasses the complexities of an OS,
allowing for cycle-accurate measurement of the extension's performance using the RISC-V
mcycle and minstret counters.

\subsection{The AES implementation}
\label{subsec:aes_impl}
The AES extension focuses on accelerating the SubBytes and MixColumns transformations.
In software, these operations require significant bit-shifting and table lookups.
The hardware implementation uses the optimized GF$(2^8)$ operations implemented on the CXU.
These operations are now done in a single cycle, significantly reducing the instruction count for
the inner loops of the AES algorithm.

\subsection{The GZIP implementation}
\label{subsec:gzip_impl}
The GZIP extension targets the DEFLATE algorithm, specifically the CRC-32 calculation
and bit-shuffling required for Huffman coding. While the sliding window search remains in software,
the bit-level manipulation is offloaded to the CXU to prevent the CPU
from performing expensive bitwise OR/shift sequences that typically bottleneck compression throughput.

\subsection{TensorFlow Lite Micro MAC speedup}
\label{subsec:tflite_impl}

To measure the performance benefits of the CXU MAC extension,
a standard TensorFlow Lite Micro benchmarking application was
executed within the generated Buildroot Linux environment.
Only one type of layer, a linear fully connected layer, was tested for speedup.

Though the initial idea of this test was to integrate a full
network into the FPGA fabric, so as to test running a whole network in
one instruction, the FPGA fabric would not fit a full network on it ---
such a task would better fit an approach of precomputing values to
read-only memory.

Due to the problem of having the state be limited by slow
writing to the CSR (see ~\autoref{sec:limitations_design}),
the most fitting test to demonstrate this implementation was the
multiply-accumulate extension on tensorflow lite.
% O.F. Надайте тут обгрунтування -- ми про це багато дискутували :=)

The benchmark utilized a simple untrained
and quantized model to evaluate TinyML performance.

The execution latency was measured for both the baseline software implementation
(using standard RV32IM instructions) and the CXU-accelerated implementation.

\section{Validation through Case Studies}
\label{sec:case_studies}

To validate the effectiveness and flexibility of the CXU sandbox, three case-study
extensions were implemented and benchmarked, each paired with a software reference
implementation for direct comparison.

The \textbf{AES extension} offloads the computationally expensive finite-field
operations in $GF(2^8)$ --- specifically the SubBytes and MixColumns transformations ---
to dedicated combinational hardware logic.
Rather than executing a multi-instruction software loop of 3--4 arithmetic operations
per $GF(2^8)$ element, the hardware extension computes the finite-field arithmetic in a
single CPU cycle, as described in Section~\ref{subsec:aes_cxu}.

The \textbf{GZIP extension} targets the bit-stream manipulation operations in the
DEFLATE decompression inner loop.
Four custom instructions --- bit reversal, mask generation, Huffman index extraction,
and LZ77 address calculation --- replace sequences of ALU operations with single-cycle
combinational hardware, as described in Section~\ref{subsec:deflate_cxu}.

The \textbf{TFLite Micro MAC extension} exposes a fused multiply-accumulate operation
as a single-cycle CXU instruction, targeting the dot-product inner loops of
quantised neural network inference, as described in Section~\ref{subsec:mac}.

These case studies demonstrate how extensions of varying complexity --- from simple
bit-manipulation primitives to algebraic field arithmetic --- can be hosted and evaluated
within the CXU Playground.

\section{Testing and Benchmarking Methodology}
\label{sec:testing}

The benchmarking strategy adopted in this work is twofold, corresponding to the two
execution environments supported by the playground.

In the bare-metal environment, each CXU extension was first simulated in a
Verilator-based testbench to verify logical correctness before synthesis.
Bare-metal microbenchmarks were then executed on the FPGA to measure the exact cycle
latency of the custom instructions and verify pipeline synchronisation, using the
RISC-V \texttt{mcycle} and \texttt{minstret} hardware performance counters for
cycle-accurate measurement.

In the Linux environment, after synthesis and boot, the same workloads were timed using
hardware performance counters and standard Linux profiling tools to compare the
execution time of the software reference implementation against the CXU-accelerated
version.
For each extension, both variants were compiled for the RV32IMAFC target using the same
cross-compilation toolchain and optimisation flags --- \texttt{-O3} for bare-metal
targets and \texttt{-O3} within the Buildroot Linux environment --- ensuring a fair comparison.
% O.F. Так, стоп, а чого саме так буде чесно?!
% P. M. Бо я містайпнув, і там і там мало бути O3 ((
Execution time was measured at multiple input sizes to assess scalability and to
distinguish fixed invocation overhead from per-byte computation savings.

\section{Data Analysis}
\label{sec:data_analysis}

Performance analysis consisted of comparing the cycle counts and execution times
recorded for the software-only reference implementations against those of the
CXU-accelerated variants at the same inputs.
Where the measured speedup fell below theoretical estimates, profiling data was used to
identify the source of overhead, such as CXU invocation cost, pipeline stall cycles for
multi-cycle operations, or context-switching overhead in the Linux environment.
The results of this analysis are presented in Chapter~\ref{ch:experiments_results}.

\section{Target Hardware: Digilent Arty S7-50}
\label{sec:target_hardware}

All synthesis, implementation, and benchmarking experiments in this work were carried
out on the \textbf{Digilent Arty S7-50} development board~\cite{DigilentArtyS7},
which hosts a Xilinx Spartan-7 XC7S50 FPGA.

The Spartan-7 XC7S50 provides 52{,}160 logic cells organised as Look-Up Tables (LUTs)
and flip-flops, 2{,}700\,Kb of distributed block RAM, and 120 DSP48E1 slices.

The board itself exposes 256\,MB of DDR3L SDRAM (accessed via LiteDRAM and the
A7DDRPHY PHY), a USB-UART bridge for console access, and a 10/100 Ethernet PHY
for network-boot and file transfer.
These peripherals are sufficient to boot a full Linux image and to transfer benchmark
payloads of arbitrary size from a host machine.

The system clock is derived from the on-board 100\,MHz oscillator; the LiteX SoC
builder instantiates a \texttt{PLLE2\_ADV} PLL to generate the CPU and peripheral
clocks.
In the baseline configuration used for all experiments the CPU runs at 50\,MHz,
such that it leaves timing margin for the CXU combinational
paths without requiring per-extension timing closure tuning.

\section{Limitations}
\label{sec:limitations_design}

Several limitations of the current implementation affect the practical utility of
certain CXU features and must be taken into account when interpreting the benchmarking
results.

\subsection{CXU CSR state IO /switch latency}
\label{subsec:cxu_csr_latency}

The most significant limitation concerns the \textbf{cxidx}/\textbf{cxdata}
CSRs.
In the LiteX SoC implementation, these registers are accessed through memory-mapped I/O
over the Wishbone bus rather than as native pipeline registers.
Each CSR read or write therefore incurs the full latency of a Wishbone bus transaction,
which is substantially higher than the single-cycle register access assumed by the CXU
specification.
In practice, the shared-state mechanism in this implementation is best treated as a
one-time configuration interface, and extensions should be designed to minimize
intermediate state transfers through CSRs.

Another limitation concerns extension switching.
Because selecting the active extension requires a \textbf{cxidx} CSR write, each
switch between extensions incurs the same bus-transaction latency.
Software should therefore be structured to batch all invocations of a given extension
together before switching to another, minimizing the number of \textbf{cxidx} writes
relative to productive custom instruction calls.

Together, these two limitations imply that the performance benefits of the CXU
mechanism are most pronounced for workloads that are computationally intensive per
invocation, require minimal shared state transfers, and maintain a single active
extension for extended periods.
The benchmarks presented in Chapter~\ref{ch:experiments_results} were designed with
these constraints in mind.

\subsection{CXU Capabilities in the Linux Device Tree}
\label{subsec:lim_dts}

The Linux kernel currently has no visibility into the available CXU extensions at
boot time. Extension base addresses are hard-coded into the runtime library rather
than discovered dynamically. A more robust implementation would extend the LiteX
SoC generator to emit DTS nodes describing the CXU capabilities (index range,
operand widths, and memory-access permissions), allowing the kernel to allocate
resources and expose extensions through a standard device interface.

\subsection{Clock Frequency Limitations}
\label{subsec:lim_clock}

The maximum achievable clock frequency of the SoC is primarily
limited by the longest combinational path in the design.
In entry-level FPGA implementations,
this critical path typically runs through the
Wishbone interconnect and the combinational logic inside the CXU.

As CXU extensions become more complex and introduce deeper combinational logic,
the critical path length increases.
This reduces the maximum synthesis frequency
(Fmax, the highest clock frequency at which a digital circuit can reliably operate without timing violations.)
that the FPGA toolchain can achieve. To address this limitation,
either additional pipeline registers must be inserted inside the CXU
to break long combinational chains, or the overall system clock frequency must be reduced.

This creates a trade-off between extension complexity,
latency, and achievable clock speed. Each CXU extension should
therefore be evaluated individually and characterised in
terms of its impact on timing as part of the development workflow.

% O.F. Щось я попереднє речення взагалі не зрозумів... Думаю, варто переформулювати чи пояснити

\subsection{Multi-Process CXU Access and Kernel Driver}
\label{subsec:lim_multiprocess}

The current userspace interface relies on a lightweight UIO (Userspace I/O) driver,
which is a thin wrapper aroung the CXU call instructions.
It only ensures that the call by itself is cleaned up
(the proper CXU is selected and the inputs sanitized to avoid improper calling of the CXU),
and, still, it provides a way to call the CXU outright, without the overhead of sanitizing inputs.
% O.F. Додайте тут трохи подробиць про UIO
This provides no protection against concurrent access from multiple processes and
does not save or restore CXU state during context switches. A proper kernel-level
driver is required to handle mutual exclusion and to integrate CXU state
save/restore into the task-switch path. Without this, running CXU-accelerated
applications in a preemptive multi-process workload risks state corruption.


% O.F. Чого тут дуже і дуже бракує -- опису вашого заліза! 
