\chapter{Related Work}
\label{ch:related_work}

The growing interest in domain-specific hardware acceleration on RISC-V processors
has produced a range of interface standards, development frameworks, and sandboxes,
each addressing part of the problem this thesis targets.
This chapter surveys the most relevant prior work across three dimensions: the
interface specifications that define how custom extensions attach to a RISC-V core
(Section~\ref{sec:related_riscv_extensions}), the SoC frameworks used to build
the surrounding hardware platform (Section~\ref{sec:related_fpga_frameworks}), and
the sandboxes designed to lower the barrier to prototyping custom instructions
(Section~\ref{sec:related_sandboxes}).
The chapter closes with a survey of application domains --- cryptography, compression,
and machine learning --- where prior work has demonstrated the practical value of
tightly coupled custom instructions, establishing the benchmarks against which
the extensions developed in this work are evaluated
(Section~\ref{sec:related_applications}).

\section{RISC-V Extensions For Custom Instruction Integration}
\label{sec:related_riscv_extensions}

Several interface specifications have been proposed to standardize the connection between
a RISC-V processor and custom hardware accelerators.
These specifications differ considerably in the number of extensions they support,
the degree of OS integration they define, and the constraints they impose on the
programmer, all of which directly motivated the design choices made in this work.

\subsection{CFU}
\label{subsec:cfu}

The Custom Function Unit (CFU) specification~\cite{Gray2022CFU, Gray2025CXU, Gray2025CXUGithub}
was proposed by Jan Gray et al.\ to make custom extensions modular and reusable
across different RISC-V CPU implementations.
The motivation was straightforward: many RISC-V processors incorporate some form of
custom extension --- for instance, a SIMD multiply-accumulate unit --- yet no
cross-implementation interoperability exists, resulting in duplicated effort whenever
the same acceleration is ported to a new core.
The CFU addresses this by defining a common logic interface to which a single custom
extension can be attached, along with timing, logical, and combinatorial constraints.
The full interface specification is examined in Chapter~\ref{ch:theoretical_background}.

Despite its value in establishing the concept of composable extensions, the CFU has
important limitations.
Only one extension may be connected per hardware thread; the extension may not access
memory, and even when programming in a higher-level language, the user must invoke a
\textbf{custom-0|1|2|3} assembly instruction directly~\cite{Gray2022CFU}.
These constraints make the CFU impractical for workloads that require multiple distinct
accelerators or expect a clean programming model.

\subsection{CXU}
\label{subsec:cxu}

The CXU specification~\cite{Gray2025CXU} is the successor to the CFU, designed to
address its principal limitations.
It supports up to $2^{\textbf{XLEN}}$,
where \textbf{XLEN} is the number of bits in it's general-purpose registers,
simultaneous extensions, permits extensions to
maintain a shared state, and defines the conditions under which an extension may access
memory. % O.F. Розшифруйте тут XLEN -- не зважаючи на те, що він є в списку абревіатур
Notably, the CXU specification also defines a standardized Application Binary Interface
(ABI), making it possible to invoke an extension from C without inline assembly.
A reference runtime implementation is provided, in which calls to CXU extensions are
wrapped in helper functions that select the appropriate extension via the
\texttt{cxidx} CSR, save the previous state, and clean up the context afterward.
The specification was developed with operating system support in mind and explicitly
defines how the kernel must save and restore CXU state during context switches.
As noted in Chapter~\ref{ch:introduction}, to the best of authors' knowledge, no hardware implementation of this interface existed prior to the present work, which forms the primary gap this thesis fills.
% O.F. Дуже ретельно це перевірте! 

\subsection{Core-V-XIF}
\label{subsec:corev_xif}

The \textbf{Core-V} E\textbf{X}tensible \textbf{I}nter\textbf{F}ace
\cite{OpenHW2021XIF} is an alternative RISC-V extension interface that supports
custom ISA extensions and coprocessors.
Like the composable custom extensions family, Core-V-XIF provides a bus-level interface
that any compliant CPU can expose.
Unlike the CFU and CXU, it permits custom CSRs and imposes fewer restrictions on
extension behaviour, offering greater flexibility at the cost of a less uniform
programming model.
A sandbox built on Core-V-XIF, FRANCIS-V~\cite{Egert2023FRANCISV}, is discussed further
in Section~\ref{sec:related_sandboxes}; however, because FRANCIS-V is tightly coupled
to a specific CPU implementation, extensions developed for it cannot be straightforwardly
ported to other cores, which limits the reusability that of the CXU specification was
designed to provide.

\section{Frameworks for FPGA-based SOC Development}
\label{sec:related_fpga_frameworks}

Beyond the interface specification, building a CXU-capable system requires a framework
that can assemble the CPU, memory, peripherals, and custom hardware logic into a coherent
SoC.
This section surveys the most relevant frameworks, noting for each where they fall short
of the needs of this work.

\subsection{LiteX}
\label{subsec:litex}

\textbf{LiteX}~\cite{LiteXGitHub} is an open-source, Python-based framework for
building FPGA-based SoCs.
It provides modules for bus integration, I/O, and CPU instantiation --- notably including
VexiiRiscv~\cite{VexiiRiscVGithub} --- as well as drivers for the most common FPGA
development board peripherals: DDR memory via LiteDRAM, Ethernet via LiteEth, PCIE,
SATA, SD cards, and standard serial buses (I2C, UART, SPI, USB).
The most widely used CFU development environment, the CFU Playground
\cite{CFUPlayground2022}, was built on LiteX, demonstrating its suitability as a base
for custom-extension sandboxes.
The present work adopts LiteX as its SoC foundation for the same reason: it abstracts
low-level hardware connectivity and provides a unified build system, allowing the
development effort to focus on the CXU integration itself rather than peripheral plumbing.

\subsection{F4PGA}
\label{subsec:f4pga}

\textbf{F4PGA}~\cite{F4PGA2019} provides a fully open-source FPGA toolchain targeting
multiple FPGA families.
While not a SoC framework by itself, it is commonly paired with LiteX or custom SoC
designs to enable reproducible, vendor-independent bitstream generation.
It was considered as a build target for the CXU Playground but was not adopted in the
final implementation, as the FPGA boards used in this work are adequately supported by
vendor toolchains.

\subsection{MiSoC}
\label{subsec:misoc}

\textbf{MiSoC}~\cite{MiSoC2014} is the predecessor of LiteX and introduced the concept
of Python-based SoC generation.
Although it is now deprecated, it is historically significant as the origin of the design
patterns that LiteX inherited and extended, and is mentioned here to provide context for
the LiteX architecture described in Chapter~\ref{ch:theoretical_background}.

\subsection{PULP Platform}
\label{subsec:pulp}

\textbf{PULP}~\cite{PULP2015} is an open-source RISC-V-based SoC platform designed for
energy-efficient parallel processing and accelerator integration.
Its cluster-based architecture has made it widely used in academic research on DSP,
cryptography, and accelerator-centric SoC exploration, including the machine learning
inference extensions discussed in Section~\ref{subsec:related_ml}.
PULP was not selected as a basis for this work because its multi-core cluster model is
orthogonal to the single-core, tightly coupled CXU interface that is the subject of
this thesis.

\subsection{SpinalHDL}
\label{subsec:spinalhdl}

\textbf{SpinalHDL}~\cite{SpinalHDL2018} is a hardware construction language, built on
Scala, commonly used to implement extensible SoCs.
Its plugin-based CPU architecture enables fine-grained custom instruction and pipeline
extensions, making it directly relevant to composable extension research.
VexiiRiscv~\cite{VexiiRiscVGithub} --- the CPU core used in this work --- is implemented
in SpinalHDL, and the CXU plugin described in Chapter~\ref{ch:proposed_solution} was
developed within this framework.
The SpinalHDL programming model is introduced in
Section~\ref{subsec:vexii}.

\section{Sandboxes for Simplified Development of Custom Instruction Hardware}
\label{sec:related_sandboxes}

The interface specifications and SoC frameworks described above address the infrastructure
problem, but they do not by themselves lower the barrier to writing, testing, and
benchmarking a custom extension.
Two sandboxes have been proposed for this purpose, and both directly influenced the
design of the CXU Playground developed in this work.

\subsection{FRANCIS-V}
\label{subsec:francisv}

\textbf{FRANCIS-V}~\cite{Egert2023FRANCISV} (FRAmework for iNtegrating Custom
InstructionS into RISC-V systems) is a framework aimed at making the design and
integration of custom extensions and coprocessors more tractable.
It provides an interface and methodology for connecting custom instruction hardware
to a processor using the Core-V-XIF extension, and automates simulation, compilation,
and verification workflows.
In these respects it shares the same high-level goal as the CXU Playground.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[
    box/.style={rectangle, draw, rounded corners=3pt,
                minimum width=3.2cm, minimum height=0.75cm,
                text centered, font=\small},
    wbox/.style   = {rectangle, draw, dashed, rounded corners=4pt,
                     inner sep=8pt},
    arr/.style    = {->, >=stealth, thick},
    darr/.style   = {<->, >=stealth, thick, gray},
    node distance = 0.55cm and 1.3cm
]

\node[box, fill=purple!15]               (src)  {Custom Instr.\ RTL};
\node[box, fill=purple!10, below=of src] (gen)  {Code Generator};
\node[box, fill=purple!10, below=of gen] (drv)  {Driver / Wrapper};

\node[box, fill=teal!15,   right=2.8cm of src]  (xif)  {Core-V XIF Interface};
\node[box, fill=teal!10,   below=of xif]         (cpu)  {RISC-V CPU Core};
\node[box, fill=teal!10,   below=of cpu]         (sim)  {Simulation / Verilator};

\draw[arr] (src) -- (gen);
\draw[arr] (gen) -- (drv);

\draw[arr] (xif) -- (cpu);
\draw[arr] (cpu) -- (sim);

\draw[darr] (src)  -- (xif)  node[midway, above, font=\scriptsize]{};
\draw[darr] (drv)  -- (sim)  node[midway, below, font=\scriptsize]{};

\node[wbox, fit=(src)(gen)(drv), label={[font=\scriptsize]above:FRANCIS-V Tooling}] {};
\node[wbox, fit=(xif)(cpu)(sim), label={[font=\scriptsize]above:Target CPU Environment}] {};

\end{tikzpicture}
\caption{FRANCIS-V framework overview.}
% O.F. Вибачте мою параною, але чи картинка Вашого авторства? (Бачу, що Ви її тут будували, але мало що :=)
\label{fig:francisv}
\end{figure}

However, FRANCIS-V is tightly coupled to the Core-V-XIF interface and to the specific
CPU implementations that expose it.
Extensions developed within FRANCIS-V cannot be straightforwardly ported to other
RISC-V cores, which conflicts with the portability goal of the composable extensions
family.
The CXU Playground targets the CXU interface instead, which is explicitly designed
for cross-implementation reuse.

\subsection{CFU Playground}
\label{subsec:cfu_playground}

The CFU Playground~\cite{CFUPlayground2022} is the closest prior work to the system
developed in this thesis.
It is a full-stack, open-source environment for the rapid design and evaluation of
Custom Function Units on FPGA-based RISC-V soft CPUs, built on the LiteX framework
and originally proposed for Tiny Machine Learning (tinyML) acceleration.
The playground provides a ready-made SoC environment in which users can implement
specialised hardware logic and write software to invoke their extensions.

\begin{figure}[htb]
\centering
% O.F. !!! Оті фігурні дужки розверніть в другу сторону
\begin{tikzpicture}[
    box/.style={rectangle, draw, rounded corners=3pt,
                minimum width=3.2cm, minimum height=0.75cm,
                text centered, font=\small},
    wide/.style={rectangle, draw, rounded corners=3pt,
                minimum width=6.8cm, minimum height=0.75cm,
                text centered, font=\small},
    wbox/.style={rectangle, draw, dashed, rounded corners=4pt,
                inner sep=8pt},
    arr/.style={->, >=stealth, thick},
    darr/.style={<->, >=stealth, thick},
    node distance=0cm and 0cm
]
\node[wide, fill=blue!15] (user)  {User C / C++ Application};
\node[wide, fill=blue!10, below=of user] (cfu) {CFU Library (custom-0 intrinsics)};
\node[box, fill=green!15, below=of cfu, xshift=-2cm] (litex) {LiteX SoC (VexRiscv)};
\node[box, fill=green!10, right=of litex] (hw) {CFU Hardware (RTL)};
\node[wide, fill=orange!15, below=of litex, xshift=2cm] (fpga) {FPGA Bitstream};

\draw[decorate, decoration={brace, mirror, amplitude=6pt}, thick]
    ([xshift=-2.0cm]user.north west) -- ([xshift=-2.0cm]cfu.south west)
    node[midway, left=8pt, font=\scriptsize] {Software};
\draw[decorate, decoration={brace, mirror,  amplitude=6pt}, thick]
    ([xshift=-1.54cm]litex.north west) -- ([xshift=-2.0cm]fpga.south west)
    node[midway, left=8pt, font=\scriptsize] {Hardware};

\end{tikzpicture}
\caption{CFU Playground full-stack architecture.}
\label{fig:cfu_playground_arch}
\end{figure}

The CFU Playground is built on the initial version of the composable custom extensions
specification, which restricts each thread to a single extension and requires assembly-level
\textbf{custom-0|1|2|3} opcodes even in higher-level language contexts
\cite{Gray2022CFU, CFUPlayground2022}.
It does not support OS-level benchmarking,
nor multi-process sharing (unless done cooperatively, e. g. having a semaphore for the extension),
as it runs bare-metal software only.
% O.F. Бенчмаркінг то таке, чи підтримує переключення процесів? Хоча б на рівні як UART, блокуванням унікального ресурсу? Не пам'ятаю, але краще уточнічніть тут
These restrictions --- a single extension per thread, no memory access, no Linux support ---
were precisely those that motivated the CXU specification.
The CXU Playground developed in this work can therefore be understood as the natural
successor to the CFU Playground: it adopts the same LiteX-based architecture but
realises the more capable CXU interface and adds full Linux support, enabling the
system-level evaluation that the CFU Playground cannot provide.

\section{Applications of Custom Instruction Extensions}
\label{sec:related_applications}

The interface standards and sandboxes above exist to serve a practical need: a wide
range of computationally intensive algorithms benefit substantially from narrowly
scoped hardware acceleration.
This section surveys the application domains most relevant to the case studies
developed in this work --- cryptography, data compression, and machine learning --- and
identifies the results against which the CXU extensions are compared in
Chapter~\ref{ch:experiments_results}.

\subsection{AES and Symmetric Cryptography}
\label{subsec:related_aes}

AES~\cite{NISTAES} is based on a substitution-permutation network whose inner
operations: SubBytes (an S-box lookup over $\text{GF}(2^8)$) and MixColumns
(matrix multiplication over $\text{GF}(2^8)$) --- map naturally onto
combinational hardware.
Cui and Balasch~\cite{Cui2023MaskingAES} demonstrate this concretely in the context
of masked AES, introducing constant-time finite-field instructions that simultaneously
improve performance and resistance to side-channel attacks.
Their result is particularly relevant here because it shows that a small number of
carefully scoped ISA extensions --- rather than a monolithic AES coprocessor --- can
yield substantial gains while preserving software flexibility.

Primas et al.~\cite{Primas2024NLUV} take a related approach with the NLU-V
extension, fusing S-box substitution and linear diffusion into a single hardware
block to minimise intermediate data movement for both AES and PRESENT.
Both works support the design choice made in Chapter~\ref{ch:proposed_solution}
to expose $\text{GF}(2^8)$ arithmetic as a primitive CXU instruction rather than
implementing a full AES round in hardware.

\subsection{Post-Quantum Cryptography}
\label{subsec:related_pqc}

Classical public-key cryptosystems are vulnerable to Shor's algorithm on sufficiently
powerful quantum computers, motivating the NIST post-quantum standardisation effort
\cite{Vranken2023UniformMul, Levi2020RISQV}.
Lattice-based schemes such as CRYSTALS-Kyber and NewHope rely on polynomial arithmetic,
particularly the Number Theoretic Transform (NTT), whose butterfly operations and
modular reductions are expensive on a standard ALU.

Sozbilir et al.~\cite{Sozbilir2020Kyber} prototype NTT-specific ISA extensions on
VexRiscv and report up to an 85\% speedup for Kyber and NewHope, demonstrating
that the class of processor used in this work is a viable target for post-quantum
acceleration.
Levi et al.~\cite{Levi2020RISQV} extend this further with RISQ-V, achieving a
$9.6\times$ speedup for Kyber using 29 tightly coupled custom instructions.
Vranken et al.~\cite{Vranken2023UniformMul} argue for a uniform set of
multiply-accumulate and modular-reduction primitives capable of serving both
classical and post-quantum schemes, which directly foreshadows the approach taken
for the MAC extension described in Section~\ref{sec:case_studies}.
These results collectively motivate extending the CXU Playground to post-quantum
workloads as a natural direction for future work, discussed further in
Section~\ref{sec:future_work}.

\subsection{Hashing}
\label{subsec:related_hashing}

Alawadi et al.~\cite{Alawadi2025SHA3} implement the full Keccak-f permutation as a
single custom RISC-V instruction, reporting substantial reductions in instruction count
and cycle time for SHA-3.
Their area-performance trade-off analysis provides a useful reference for understanding
the cost of integrating high-complexity primitives into the CXU pipeline.

\subsection{Data Compression}
\label{subsec:related_compression}

GZIP/DEFLATE combines LZ77 back-reference encoding with Huffman entropy coding
\cite{DEFLATESpec}.
The performance bottleneck in decompression lies in the inner loop, which requires
repeated bit-field extractions, variable-width mask generation, and modular address
arithmetic --- all operations that reduce to a handful of hardware gates but require
several ALU instructions in software.
The custom instructions designed in Section~\ref{subsec:deflate_cxu} and evaluated
in Chapter~\ref{ch:experiments_results} target exactly this hot path.
To the best of the authors' knowledge, no prior work has applied the CXU or CFU
mechanism to GZIP acceleration, making this a novel case study.

\subsection{Machine Learning Inference}
\label{subsec:related_ml}

Neural network inference on microcontrollers is dominated by Multiply-Accumulate (MAC)
operations in convolutional and fully connected layers.
TensorFlow Lite Micro~\cite{David2021TFLiteMicro} provides portable INT8-quantised
inference kernels for this class of device.
The CMSIS-NN library~\cite{Lai2018CMSISNN} shows that ARM DSP SIMD MAC instructions
yield up to a $4.6\times$ improvement over scalar code; the PULP platform
\cite{PULP2015} achieves comparable gains on RISC-V with hardware-loop and
dot-product extensions.
The MAC extension evaluated in Chapter~\ref{ch:experiments_results} follows the same
principle, exposing a fused multiply-accumulate operation as a single-cycle CXU
instruction.
Oliveira et al.~\cite{Oliveira2022MiniFloat} push this further with float8
dot-product accumulation for low-precision training, demonstrating that the same
CXU-style integration pattern scales to more exotic arithmetic formats.

\subsection{Linux on LiteX}
\label{subsec:related_linux_litex}

Running benchmarks inside a full operating system yields performance numbers that
reflect real-world deployment conditions rather than isolated microbenchmarks.
The Linux-on-LiteX-VexRiscv project~\cite{LinuxOnLiteXGitHub} establishes the
feasibility of booting Linux on a soft RISC-V core in a LiteX SoC, and documents
the required hardware configuration: MMU support, atomic instructions, and supervisor
privilege mode.
This work adopts the same configuration and extends it with the CXU hardware,
as described in Section~\ref{sec:sandbox_env}.

While prior work has demonstrated the value of custom instructions
for AES, post-quantum cryptography, hashing, machine learning, and (to a lesser
extent) compression, none has done so within the CXU framework, and none has
provided a unified sandbox environment that supports multiple extensions under Linux.
The present work addresses both gaps simultaneously.
% O.F. Ретельно перевірте, чи це правда! 
