\chapter{Theoretical Background}
\label{ch:theoretical_background}

This chapter provides the technical background required to understand the design decisions
made in Chapters~\ref{ch:proposed_solution} and~\ref{ch:experiments_results}.
It begins with the RISC-V ISA and the VexiiRiscv processor core that serves as the
execution target (Section~\ref{sec:riscv_vexii}), then describes the LiteX SoC framework
used to assemble the surrounding hardware platform
(Sections~\ref{sec:litex_ecosystem}--\ref{sec:buildroot}).
Section~\ref{sec:cxu_mechanism} examines the CXU coupling architecture and the
VexiiRiscv plugin interface through which the extension is implemented.
Section~\ref{sec:linux_litex} covers the requirements and boot flow for running Linux on
the generated SoC.
The chapter closes with the algorithmic background for the three case-study workloads:
AES encryption (Section~\ref{sec:aes}), GZIP/DEFLATE compression
(Section~\ref{sec:gzip}), and TensorFlow Lite Micro inference
(Section~\ref{sec:tflite}).

\section{The RISC-V Architecture and VexiiRiscv Core}
\label{sec:riscv_vexii}

\subsection{Overview of the RISC-V ISA}
\label{subsec:riscv_isa}

The RISC-V instruction set architecture (ISA) \cite{RISC-VInternational2025} is an open
standard designed with simplicity, extensibility, and long-term sustainability in mind.
Unlike proprietary architectures such as x86 or ARM, RISC-V is developed under an
open governance model and can be freely implemented without licensing constraints.

A central design principle of RISC-V is modularity.
The base ISA provides only essential functionality, while optional extensions add
specialized capabilities.
Common extension subsets include integer multiplication (\textbf{IM}),
atomic operations (\textbf{A}),
single- and double-precision floating-point arithmetic (\textbf{F/D}),
and compressed 16-bit instructions (\textbf{C}).
Typical Linux-capable systems use configurations such as RV64IMAC \cite{Rakhmatov2023Survey};
this work uses RV32IMAFC, as required by the Linux-on-LiteX platform
\cite{LinuxOnLiteXGitHub}, specifically for the integer and atomic instructions
used for the synchronization primitives created in linux, as well as
floating point operations to simplify compiling some libraries used in this work.
% O.F. Тут може варто пару слів -- а чому вимагає саме так

Notably for this thesis, RISC-V reserves opcode space for custom extensions.
These custom instruction fields allow designers to introduce application-specific
accelerators directly into the processor pipeline without conflicting with standard
instructions, and form the opcode basis for the CXU interface described in
Section~\ref{sec:cxu_mechanism}.

\subsection{The VexiiRiscv Implementation}
\label{subsec:vexii}

VexiiRiscv \cite{VexiiRiscVGithub} is a configurable RISC-V core implemented in
SpinalHDL \cite{SpinalHDL2018}, a Scala-based hardware description language.
Unlike traditional Verilog or VHDL, SpinalHDL supports generative and highly
parameterized hardware construction, enabling scalable design-space exploration,
easier feature composition, and cleaner abstraction boundaries.

The processor follows a classical pipelined architecture with up to five stages:
\textbf{Fetch}, \textbf{Decode}, \textbf{Execute}, \textbf{Memory}, and \textbf{Write-back},
where Fetch and Execute may be unified, and Memory and Write-back may also be unified.
One of the distinguishing characteristics of VexiiRiscv is its plugin-based architecture:
% O.F. Може варто пару слів більше про те, як ті плагіни функціонують
major features --- including memory management units (MMU), caches, branch predictors,
and custom execution units --- are implemented as independent plugins that can be composed
at generation time. These plugins are created as simple logic modules, that are enabled
after some event. They are the connected to the overall logic, and triggered when necessary.
The CXU extension is implemented as an execute-stage plugin; all computation is performed by
CXU coprocessors and therefore occurs in the Execute stage of the pipeline, as described
further in Section~\ref{sec:cxu_mechanism}.

\section{The LiteX SoC Ecosystem}
\label{sec:litex_ecosystem}

Having established the processor core, this section describes the SoC framework that
surrounds it.
LiteX provides the integration layer between VexiiRiscv and the FPGA peripherals,
and its memory-mapped architecture underpins the bus transactions used by the CXU's
CSR interface.

\subsection{LiteX Framework Overview}
\label{subsec:litex_overview}

LiteX \cite{LiteXGitHub} is a Python-based SoC construction framework built on top of
Migen \cite{MiSoC2014}.
It enables automated generation of FPGA-based SoCs by integrating CPU cores,
memory controllers, interconnects, and peripherals into a cohesive hardware platform.
Within this ecosystem, LiteX acts as the integration layer between the VexiiRiscv CPU
core and FPGA peripherals such as UART, Ethernet \cite{LiteEthGithub}, timers, and
external memory controllers \cite{LiteDramGithub, LitePcieGithub}.
The framework abstracts low-level hardware connectivity, automatically generates HDL,
memory maps, CSR definitions, and software headers, and provides a unified build system
for generating bitstreams and associated firmware.

\subsection{Memory-Mapped Architecture}
\label{subsec:memory_mapped}

LiteX uses memory-mapped I/O as its primary software-hardware interface: peripherals
are accessed by reading or writing specific addresses, such as:

\begin{verbatim}
#define UART_BASE 0xF0001000
\end{verbatim}

Software interacts with hardware exclusively by manipulating these addresses.
Figure~\ref{fig:litex_soc} illustrates the resulting hardware architecture of the
generated SoC, showing the VexiiRiscv CPU (with CXU Plugin) and the LiteEth DMA
controller as bus masters, connected via the Wishbone interconnect to memory and
peripheral slaves.

\begin{figure}[htb]
    \centering
    \begin{tikzpicture}[
        node distance=1.5cm and 0.5cm,
        master/.style={draw, rectangle, minimum width=3.5cm, minimum height=1.5cm, align=center, fill=red!10, rounded corners, font=\bfseries},
        slave/.style={draw, rectangle, minimum width=2.5cm, minimum height=1.5cm, align=center, fill=blue!10, rounded corners},
        bus/.style={draw, rectangle, minimum width=15cm, minimum height=1cm, align=center, fill=gray!20, font=\bfseries\large},
        arrow/.style={-Latex, thick},
        bidir/.style={Latex-Latex, thick},
        group/.style={draw, dashed, thick, fill=gray!5, rounded corners, inner sep=0.5cm}
        box/.style={rectangle, draw, rounded corners=3pt,
            minimum width=3.2cm, minimum height=0.75cm,
            text centered, font=\small},
    ]

    \node[bus] (wishbone) {Wishbone Interconnect (32-bit)};

    \node[master, above=1.5cm of wishbone, xshift=-4cm] (cpu) {VexiiRiscv CPU \\ \small + CXU Plugin};
    \node[master, above=1.5cm of wishbone, xshift=4cm, fill=orange!10] (dma) {LiteEth DMA \\ \small (Network Master)};

    \node[slave, below=1.5cm of wishbone, xshift=-6cm] (rom) {BootROM / \\ \small SPI Flash};
    \node[slave, below=1.5cm of wishbone, xshift=-3cm] (litedram) {LiteDRAM \\ \small (Main Memory)};
    \node[slave, below=1.5cm of wishbone, xshift=0cm] (liteuart) {LiteUART \\ \small (Serial Console)};
    \node[slave, below=1.5cm of wishbone, xshift=3cm] (liteeth) {LiteEth MAC \\ \small (Network Slave)};
    \node[slave, below=1.5cm of wishbone, xshift=6cm] (timer) {Timer / \\ \small Core CSRs};

    \draw[bidir] (cpu.south) -- (cpu.south |- wishbone.north) node[midway, left, font=\small] {Instr / Data};
    \draw[bidir] (dma.south) -- (dma.south |- wishbone.north) node[midway, right, font=\small] {RX / TX};

    \draw[bidir] (rom.north |- wishbone.south) -- (rom.north);
    \draw[bidir] (litedram.north |- wishbone.south) -- (litedram.north);
    \draw[bidir] (liteuart.north |- wishbone.south) -- (liteuart.north);
    \draw[bidir] (liteeth.north |- wishbone.south) -- (liteeth.north);
    \draw[bidir] (timer.north |- wishbone.south) -- (timer.north);

    \end{tikzpicture}
    \caption{Hardware architecture of the generated LiteX SoC.}
    \label{fig:litex_soc}
\end{figure}

\subsection{Interconnect and Control}
\label{subsec:interconnect}

LiteX uses the Wishbone bus as its on-chip interconnect standard.
Wishbone is a simple, flexible, synchronous bus protocol that supports
master-slave communication between cores and peripherals, enabling memory-mapped I/O
and simplifying peripheral integration \cite{LiteXGitHub}.


Control and Status Registers (CSRs) provide the primary mechanism for software-hardware
communication. They are used for overal communication with the CPU
(as, for example, for reading  a timer register), and for interaction with the CXU state.
% O.F.  Тут може варто уточнити, з якою хардою комунікації -- тут плутає memory mapper природа LiteX-комунікації і спосіб комунікації з CXU. Чи cxidx тут і є memory mapped? 
% O.F. Знайдений розділ limitations трохи прояснив -- але, думаю, врато і тут
CSRs are memory-mapped registers that allow software to configure hardware blocks,
query status information, and trigger operations~\cite{LiteDramGithub, LiteXGitHub}.
This abstraction is essential for interacting with accelerators and managing runtime
behavior; in the CXU implementation, the \texttt{cxidx} and \texttt{cxdata} registers
through which software selects the active extension are CSRs of this kind.
An important consequence --- discussed as a limitation in
Section~\ref{sec:limitations_design} --- is that CSR access incurs a full Wishbone bus
transaction, which is substantially more expensive than the register-file access assumed
by the CXU specification.

\subsection{Memory Management}
\label{subsec:memory_mgmt}

LiteX integrates external memory controllers through LiteDRAM \cite{LiteDramGithub},
which supports SDRAM and DDR memory devices.
LiteDRAM abstracts memory timing, initialization, and refresh management, exposing a
consistent interface to the system bus.
The SoC memory map defines address regions for ROM, SRAM, DRAM, CSRs, and peripherals.
Proper memory layout is particularly important for Linux-based systems, where kernel
placement, device tree configuration, and bootloader operation all depend on predictable
address organization.

\section{Software Development for LiteX-Based SoCs}
\label{sec:sw_dev}

Developing software for a LiteX-based SoC requires cross-compilation against the
target RISC-V architecture and careful linking against the target memory map.

\subsection{Cross-Compilation}
\label{subsec:cross_compile}

Software is compiled on a host machine using a RISC-V cross-compilation toolchain.
The resulting ELF binary is then loaded into on-chip memory (SRAM, DRAM, or SPI Flash),
as used in \cite{CXUPlayground}.

\subsection{Linker Scripts}
\label{subsec:linker_scripts}

LiteX SoCs define a memory map that must be respected during linking.
A linker script specifies the instruction memory region, data memory region, and stack
and heap placement.
Correct linking ensures the program executes from valid memory and interacts properly
with memory-mapped peripherals.

\subsection{Bare-Metal vs Linux-Based Systems}
\label{subsec:baremetal_linux}

The LiteX platform supports bare-metal, RTOS-based, and Linux operating systems.
Bare-metal software directly accesses memory-mapped registers and can use hardware
performance counters (such as the RISC-V \texttt{mcycle} counter) for cycle-accurate
measurement.
Linux-based systems require drivers and standard system calls, but provide a more
realistic deployment environment; the benchmarks in
Chapter~\ref{ch:experiments_results} uses both modes to separate hardware-only
latency from system-level overhead.

\section{Buildroot}
\label{sec:buildroot}

To run and benchmark CXU applications under Linux, the system requires a complete
embedded Linux image.
\textbf{Buildroot} is an embedded Linux build system that automates the generation of
complete root filesystems, cross-toolchains, and kernel images for embedded targets.
Its workflow consists of selecting the target architecture, enabling desired packages,
and building a complete system image; the output typically includes the kernel image,
a Device Tree Blob (DTB), and a filesystem image.
Buildroot is commonly used with LiteX SoCs to produce bootable Linux images for
FPGA-based systems \cite{CXUPlayground}, and was adopted in this work for exactly
that purpose.
The integration of Buildroot into the CXU Playground build process is described in
Section~\ref{sec:sandbox_env}.

\section{Custom Extension Units (CXU) Mechanism}
\label{sec:cxu_mechanism}

This section describes the architectural principles behind the CXU interface and the
specific way in which it is implemented within the VexiiRiscv pipeline.
Understanding these details is a prerequisite for the implementation discussion in
Chapter~\ref{ch:proposed_solution}.

\subsection{Coupling Architecture}
\label{subsec:coupling}

Hardware accelerators can be integrated into a system using either tightly coupled or
loosely coupled approaches \cite{Gray2022CFU, Gray2025CXU}.
Tightly coupled accelerators are directly integrated into the CPU pipeline and operate as
execution units, and are invoked through custom instructions.
This approach offers low latency and direct register-file access, but requires careful
hazard and pipeline management.
Loosely coupled accelerators are attached via system buses or DMA engines and operate
as independent hardware blocks, often accessed through memory-mapped interfaces.
While simpler to integrate, they typically incur higher invocation latency.

The CXU mechanism adopts the tightly coupled strategy, enabling instruction-level
acceleration with minimal software overhead.
Figure~\ref{fig:cxu_cpu} illustrates the logical interface between the CPU and a CXU
accelerator: source operands \texttt{rs1} and \texttt{rs2} are forwarded from the
register file to the accelerator along with the opcode and \texttt{cxidx}, and the
result is written back to the destination register \texttt{rd} in the same pipeline
cycle (or after a stall for multi-cycle operations).

\begin{figure}[htb]
    \centering
    \begin{tikzpicture}[
        block/.style={draw, rectangle, minimum width=3cm, minimum height=4.5cm, align=center},
        arrow/.style={-Latex, thick},
        box/.style={rectangle, draw, rounded corners=3pt,
            minimum width=3.2cm, minimum height=0.75cm,
            text centered, font=\small}
    ]
        \node[block, fill=blue!10] (cpu) {VexiiRiscv CPU};
        \node[block, fill=green!10, right=4cm of cpu] (cxu) {CXU Accelerator};

        \draw[arrow] ([yshift=1.5cm]cpu.east) -- node[above, font=\small] {rs1 (32-bit)} ([yshift=1.5cm]cxu.west);
        \draw[arrow] ([yshift=0.5cm]cpu.east) -- node[above, font=\small] {rs2 (32-bit)} ([yshift=0.5cm]cxu.west);
        \draw[arrow] ([yshift=-0.5cm]cpu.east) -- node[above, font=\small] {cxidx / opcode} ([yshift=-0.5cm]cxu.west);

        \draw[arrow] ([yshift=-1.5cm]cxu.west) -- node[above, font=\small] {rd (32-bit)} ([yshift=-1.5cm]cpu.east);

        \draw[arrow, dashed] ([yshift=2.2cm]cpu.east) -- node[above, font=\small\itshape] {Valid / Cmd Enable} ([yshift=2.2cm]cxu.west);
        \draw[arrow, dashed] ([yshift=-2.2cm]cxu.west) -- node[above, font=\small\itshape] {Ready / Pipeline Stall} ([yshift=-2.2cm]cpu.east);
    \end{tikzpicture}
    \caption{Logical interface between the VexiiRiscv CPU and a CXU accelerator.}
    \label{fig:cxu_cpu}
\end{figure}

\subsection{VexiiRiscv CXU User Plugin Interface}
\label{subsec:cxu_plugin_iface}

The VexiiRiscv plugin interface enables custom instruction integration at multiple
stages of the execution pipeline. It allows for execution of ablocking
instruction that freezes the pipeline on the execution stage,
for non-blocking extenisons that may not freeze the pipeline at all,
or for extensions that may freeze the pipeline on the write-back stage.
% O.F. Зізнаюся, не впевнений, чи розумію це речення...
Within the CXU plugin, instruction handling proceeds through the following steps:

\begin{enumerate}
    \item CXU state is updated through the \texttt{cxidx} and \texttt{cxdata} CSRs.
    \item The CXU custom command is called with \texttt{rs1} and \texttt{rs2} as inputs and \texttt{rd} as output.
    \item Source operands are read from registers \texttt{rs1} and \texttt{rs2}.
    \item The CXU processes the input state and registers and returns the desired output.
    \item The result is written back to the destination register \texttt{rd}.
    \item Updated CXU state may be read back through the \texttt{cxidx} and \texttt{cxdata} CSRs.
\end{enumerate}

\subsubsection{Instruction Decoding}

Specific opcode patterns reserved for custom instructions are mapped to the CXU during
the decode stage \cite{SpinalHDL2018}.
Matching instructions are redirected to the custom execution logic by SpinalHDL's
built-in extension mechanism, after which the CXU connector extension forwards the
decoded command --- including the operands and the active extension index --- to the
coprocessor.

Multi-cycle CXU operations require pipeline stall mechanisms to preserve correctness.
The plugin must signal readiness to the CPU and manage structural, data, and control
hazards to ensure correct execution ordering.
The implementation of these stall signals within the VexiiRiscv pipeline is described
in Chapter~\ref{ch:proposed_solution}.

\section{Linux Support on LiteX}
\label{sec:linux_litex}

Running the CXU Playground under Linux requires specific hardware capabilities from
the SoC and a well-defined boot sequence.
This section describes both, establishing the prerequisites for the Buildroot integration
discussed in Section~\ref{sec:sandbox_env}.

\subsection{Hardware Requirements}
\label{subsec:hw_requirements}

Running Linux on a RISC-V SoC requires three hardware capabilities
\cite{LinuxOnLiteXGitHub}: a Memory Management Unit (MMU) for virtual memory support,
atomic instruction support (RV32A) for synchronisation primitives, and supervisor
privilege mode for kernel-user separation.
These features are enabled through the appropriate VexiiRiscv plugins and LiteX SoC
configuration parameters during bitstream generation.

\subsection{The Boot Process}
\label{subsec:boot_process}

The Linux boot flow on a LiteX SoC proceeds through multiple stages, as illustrated
in Figure~\ref{fig:boot_process}.
Following a hardware reset, the LiteX BIOS performs hardware
initialization and memory flashing.
OpenSBI \cite{OpenSBIGithub} then provides the Supervisor Binary Interface (SBI),
acting as a firmware layer between the machine mode and the operating system, handling
low-level initialization and exposing standardized services to the kernel.
A secondary bootloader, such as U-Boot \cite{UBootWebsite}, loads the Linux kernel image
and an optional initial ramdisk into memory, after which the Linux kernel initialises
drivers, the memory management subsystem, and user-space services.

\begin{figure}[htb]
    \centering
    \begin{tikzpicture}[
        node distance=0.8cm,
        box/.style={rectangle, draw, rounded corners=3pt,
            minimum width=3.2cm, minimum height=0.75cm, align=center,
            text centered, font=\small}
    ]
        \node[box, fill=gray!20] (reset) {Hardware Reset};
        \node[box, fill=blue!20, below=of reset] (LiteX) {LiteX BIOS (Bootrom)\\ \small Hardware Init \& Memory Training};
        \node[box, fill=purple!20, below=of LiteX] (opensbi) {OpenSBI\\ \small Machine Mode Environment Setup};
        \node[box, fill=orange!20, below=of opensbi] (linux) {Linux Kernel\\ \small Supervisor Mode (MMU, DTB Parsing)};
        \node[box, fill=cyan!20, below=of linux] (user) {Buildroot User Space\\ \small Init Scripts \& CXU Applications};

        \draw[->] (reset) -- (LiteX);
        \draw[->] (LiteX) -- (opensbi);
        \draw[->] (opensbi) -- (linux);
        \draw[->] (linux) -- (user);
    \end{tikzpicture}
    \caption{The LiteX SoC Linux boot process.}
    \label{fig:boot_process}
\end{figure}

\subsection{Device Tree Overlay (DTS)}
\label{subsec:dts}

The device tree describes the hardware layout to the Linux kernel \cite{DeviceTreeDoc}.
LiteX generates a hardware description that is translated into a Device Tree Source
(DTS) file specifying CPU configuration, memory regions, peripheral addresses, and
interrupt mappings.
The kernel uses this information to configure drivers and manage hardware resources
dynamically.
A known limitation of the current implementation --- that CXU capabilities are not yet
described in the device tree --- is discussed in Section~\ref{subsec:lim_dts}.

\section{The AES Encryption}
\label{sec:aes}

The Advanced Encryption Standard (AES) \cite{NISTAES} is a symmetric block cipher
established by the U.S.\ National Institute of Standards and Technology and widely
used for encrypting sensitive information.
AES is based on a substitution-permutation network and operates on a 16-byte block
of data arranged in column-major order, referred to as the \textbf{State}.
The number of transformation rounds depends on the key size: 10 rounds for 128-bit
keys, 12 rounds for 192-bit keys, and 14 rounds for 256-bit keys.
Each round consists of several processing steps; inverse rounds are applied during
decryption to recover the plaintext.
The full encryption pipeline is illustrated in Figure~\ref{fig:aes_flow}.

\begin{figure}[!htb]
    \centering
    \begin{tikzpicture}[
        node distance=0.6cm and 1.5cm,
        box/.style={draw, rectangle, minimum width=3.5cm, minimum height=0.8cm, align=center, fill=blue!10, rounded corners},
        keybox/.style={draw, rectangle, minimum width=2.5cm, minimum height=0.8cm, align=center, fill=red!10, rounded corners},
        io/.style={draw, trapezium, trapezium left angle=70, trapezium right angle=110, minimum width=3cm, minimum height=0.8cm, align=center, fill=green!10},
        arrow/.style={-Latex, thick},
        group/.style={draw, rectangle, dashed, thick, inner sep=0.4cm, fill=gray!10, rounded corners}
    ]
    \node[io] (plaintext) {Plaintext (128-bit)};

    \node[box, below=1cm of plaintext] (init_addkey) {AddRoundKey};
    \node[keybox, right=of init_addkey] (key0) {Round Key 0};

    \node[box, below=1.2cm of init_addkey] (subbytes) {SubBytes};
    \node[box, below=of subbytes] (shiftrows) {ShiftRows};
    \node[box, below=of shiftrows] (mixcols) {MixColumns};
    \node[box, below=of mixcols] (main_addkey) {AddRoundKey};
    \node[keybox, right=of main_addkey] (key1_9) {Round Keys 1--9};

    \node[box, below=2cm of main_addkey] (subbytes_f) {SubBytes};
    \node[box, below=of subbytes_f] (shiftrows_f) {ShiftRows};
    \node[box, below=of shiftrows_f] (fin_addkey) {AddRoundKey};
    \node[keybox, right=of fin_addkey] (key10) {Round Key 10};

    \node[io, below=1cm of fin_addkey] (ciphertext) {Ciphertext (128-bit)};

    \begin{scope}[on background layer]
        \node[group, fit=(subbytes) (shiftrows) (mixcols) (main_addkey),
              label=left:\rotatebox{90}{\textbf{9 Standard Rounds}}] (main_loop) {};

        \node[group, fit=(subbytes_f) (shiftrows_f) (fin_addkey),
              label=left:\rotatebox{90}{\textbf{Final Round}}] (final_loop) {};
    \end{scope}

    \draw[arrow] (plaintext) -- (init_addkey);
    \draw[arrow] (key0) -- (init_addkey);
    \draw[arrow] (init_addkey) -- (subbytes);
    \draw[arrow] (subbytes) -- (shiftrows);
    \draw[arrow] (shiftrows) -- (mixcols);
    \draw[arrow] (mixcols) -- (main_addkey);
    \draw[arrow] (key1_9) -- (main_addkey);

    \draw[arrow, rounded corners=5pt] (main_addkey.south) -- ++(0,-0.4) -|
          ([xshift=-1cm]main_loop.west) |- node[pos=0.25, left] {Loop 9x}
          ([yshift=0.4cm]subbytes.north) -- (subbytes.north);

    \draw[arrow] (main_addkey) -- (subbytes_f);
    \draw[arrow] (subbytes_f) -- (shiftrows_f);
    \draw[arrow] (shiftrows_f) -- (fin_addkey);
    \draw[arrow] (key10) -- (fin_addkey);
    \draw[arrow] (fin_addkey) -- (ciphertext);

    \end{tikzpicture}
    \caption{Flowchart of the AES encryption algorithm.}
    \label{fig:aes_flow}
\end{figure}

\subsection{AES Round Operations}
\label{subsec:aes_rounds}

\subsubsection{KeyExpansion}

The KeyExpansion step derives the set of per-round subkeys from the original cipher
key using a schedule that applies S-box substitution and XOR with round constants
(Rcon), expanding a 128-bit key into eleven 128-bit round keys.

\subsubsection{AddRoundKey}

AddRoundKey XORs each byte of the State with the corresponding byte of the current
round key.
Being a bitwise operation, it is computationally inexpensive and contributes
negligibly to AES runtime on a standard processor.

\subsubsection{MixColumns}

MixColumns treats each column of the State as a four-element vector of $\mathrm{GF}(2^8)$
elements and multiplies it by the fixed circulant matrix:
$$
\begin{bmatrix}
02 & 03 & 01 & 01 \\
01 & 02 & 03 & 01 \\
01 & 01 & 02 & 03 \\
03 & 01 & 01 & 02
\end{bmatrix},
$$
where all arithmetic is performed modulo the AES reduction polynomial
$x^8 + x^4 + x^3 + x + 1$.
Each output byte requires a multiply-by-2, a multiply-by-3, and two identity terms
in $\mathrm{GF}(2^8)$, making MixColumns the dominant computational cost of a
software AES round.

\subsection{Speeding up AES with the CXU}
\label{subsec:aes_cxu}

The dominant computational cost of AES on a standard processor lies in the SubBytes
and MixColumns operations, which each require a loop of 5--6 iterations, performing
3--4 arithmetic operations in $GF(2^8)$ per iteration.
By implementing the $GF(2^8)$ arithmetic as a single-cycle combinational CXU
instruction, each loop body collapses to a single custom instruction, eliminating the
multi-instruction inner loop entirely.
The design and benchmarking of this extension are described in
Sections~\ref{subsec:aes_impl} and~\ref{subsec:aes_results} respectively.

\section{GZIP}
\label{sec:gzip}

\textbf{GZIP} (GNU Zip) \cite{GZIPWebsite} is a widely used lossless compression
format based on the \textbf{DEFLATE} algorithm \cite{DEFLATESpec}, which combines
dictionary-based compression and entropy encoding.
A GZIP file consists of a header (containing metadata, flags, and an optional
filename), a DEFLATE-compressed data stream, and a footer (containing a CRC-32
checksum and the original uncompressed size).

\subsection{The DEFLATE Algorithm}
\label{subsec:deflate}

DEFLATE combines two complementary techniques.
First, LZ77 detects repeated byte sequences and replaces them with compact
\textit{(length, distance)} back-references.
The resulting symbol stream --- comprising literal bytes, length codes, and distance
codes --- is then entropy-coded using Huffman coding and packed into a bitstream.
Figure~\ref{fig:deflate_flow} illustrates this pipeline.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[
    node distance=0.7cm and 1.5cm,
    box/.style={draw, rectangle, minimum width=3.5cm, minimum height=0.8cm, align=center, fill=blue!10, rounded corners},
    keybox/.style={draw, rectangle, minimum width=2.5cm, minimum height=0.8cm, align=center, fill=red!10, rounded corners},
    io/.style={draw, trapezium, trapezium left angle=70, trapezium right angle=110, minimum width=3cm, minimum height=0.8cm, align=center, fill=green!10},
    arrow/.style={-Latex, thick},
    group/.style={draw, rectangle, dashed, thick, inner sep=0.4cm, fill=gray!10, rounded corners}
]

\node[io] (input) {Input Data Stream};
\node[box, below=1cm of input] (lz77) {LZ77 Sliding Window\\Match Detection};
\node[box, below=of lz77] (tokens) {Generate Symbols\\(Literals, Length, Distance)};
\node[box, below=1.2cm of tokens] (freq) {Symbol Frequency\\Counting};
\node[keybox, right=of freq] (tree) {Huffman Tree\\Generation};
\node[box, below=of freq] (huff) {Huffman Encoding};
\node[box, below=of huff] (pack) {Bitstream Packing};
\node[io, below=1cm of pack] (output) {Compressed DEFLATE Stream};

\begin{scope}[on background layer]
\node[group, fit=(lz77) (tokens),
      label=left:\rotatebox{90}{\textbf{LZ77 Stage}}] (lzgroup) {};
\node[group, fit=(freq) (huff),
      label=left:\rotatebox{90}{\textbf{Huffman Coding}}] (huffgroup) {};
\end{scope}

\draw[arrow] (input) -- (lz77);
\draw[arrow] (lz77) -- (tokens);
\draw[arrow] (tokens) -- (freq);
\draw[arrow] (freq) -- (huff);
\draw[arrow] (tree) -- (huff);
\draw[arrow] (huff) -- (pack);
\draw[arrow] (pack) -- (output);

\end{tikzpicture}
\caption{Flowchart of the DEFLATE compression pipeline.}
\label{fig:deflate_flow}
\end{figure}

\subsection{LZ77 Compression}
\label{subsec:lz77}

LZ77 operates by maintaining a sliding window over previously seen data.
Repeated sequences are replaced with \textit{(length, distance)} pairs, where the
distance indicates how far back in the window the matching sequence begins.
After LZ77 processing, the data stream consists of literal bytes (0--255), length
codes representing match lengths, and distance codes representing match distances.
These symbols are not stored directly; they are entropy-encoded using Huffman coding,
as described below.

\subsection{Huffman Coding and Tree Modes}
\label{subsec:huffman}

Huffman coding assigns shorter bit patterns to more frequently occurring symbols,
reducing the average number of bits required to represent the data.
DEFLATE supports two tree modes: \textbf{Static Huffman}, which uses predefined code
tables defined by the DEFLATE specification and require no per-block header overhead,
and \textbf{Dynamic Huffman}, which computes a custom code table per block by counting
symbol frequencies and constructing an optimal tree.
Dynamic coding typically achieves better compression because it adapts to the local
symbol distribution of each block, at the cost of additional per-block computation.

The bit-manipulation operations that dominate the decompression inner loop ---
variable-width bit extraction, mask generation, and modular address arithmetic ---
are well-suited to CXU acceleration, as each reduces to a handful of logic gates but
requires several ALU instructions in software.
In contrast to AES, which relies on algebraic transformations in $GF(2^8)$, GZIP
compression primarily relies on pattern detection and entropy encoding to eliminate
statistical redundancy.

\subsection{Custom Instructions for DEFLATE Acceleration}
\label{subsec:deflate_cxu}

To accelerate the critical primitives in the DEFLATE decompression path, four custom
instructions were implemented using the CXU extension mechanism.
Each targets an operation that occurs frequently in the inner loop but is
disproportionately expensive on a standard ALU.

\subsubsection{Bit Reversal}

The \textbf{cxu\_bitrev} instruction reverses the order of bits in a 32-bit word,
producing an output where bit $i$ of the result corresponds to bit $31-i$ of the input:
$$
y_i = x_{31-i}.
$$
Bit reversal is useful for constructing lookup indices for variable-length codes and is
implemented purely combinationally by wiring input bits to reversed output positions.

\subsubsection{Bit Mask Generation}

The \textbf{cxu\_mask} instruction generates a mask containing the lowest $n$ bits set
to one:
$$
mask = (1 \ll n) - 1.
$$
Such masks are required when extracting variable-length bit fields from the compressed
DEFLATE bitstream; generating them in a single instruction reduces the number of shifts
and arithmetic operations in the decoding loop.

\subsubsection{Huffman Table Index Extraction}

The \textbf{cxu\_huft\_idx} instruction computes a masked index value:
$$
index = x \;\&\; ((1 \ll n) - 1).
$$
This operation is used during Huffman decoding to determine the lookup index into the
decoding table based on the current bit buffer.
By fusing mask generation and bit extraction into a single instruction, the number of
required operations in the decoding loop is reduced.

\subsubsection{LZ77 Back-Reference Address Calculation}

The \textbf{cxu\_copy\_addr} instruction accelerates address calculation for LZ77
back-references during decompression.
Given the current write position $w$ in the sliding window, a base distance value
$dist\_base$ from the Huffman code, and additional extension bits $extra$, the
instruction computes:

$$
addr = (w - dist\_base - extra) \bmod 2^{15}
$$

By combining subtraction and masking into a single hardware operation, the instruction
reduces the arithmetic overhead in the inner decompression loop.

\subsubsection{Implementation Characteristics}

The GZIP CXU module is entirely combinational and does not maintain internal state.
All four instructions operate on two 32-bit operands and produce a single 32-bit result
in the same pipeline cycle, with the state interface remaining unused.
This simplifies integration and ensures that the custom instructions behave identically
to standard arithmetic instructions from the perspective of the processor pipeline.
The benchmarking results for these instructions are presented in
Section~\ref{subsec:gzip_results}.

\section{TensorFlow Lite Micro}
\label{sec:tflite}

TensorFlow Lite for Microcontrollers (TFLite Micro) \cite{David2021TFLiteMicro} is a
framework designed to run machine learning models on devices with only kilobytes of
memory.
To optimize the memory footprint and execution speed on such constrained targets,
models deployed via TFLite Micro are typically quantized, converting 32-bit
floating-point weights and activations into 8-bit integers (INT8).

\subsection{Multiply-Accumulate Operations}
\label{subsec:mac}

The most computationally expensive layers in quantized inference --- Convolutional
(Conv2D) and Fully Connected (Dense) layers --- execute as a series of dot products.
The fundamental operation in each dot product is the Multiply-Accumulate (MAC):
$$
acc = acc + (weight \times activation).
$$

On a standard RISC-V RV32IM implementation, a single MAC requires fetching the
operands, executing a \textbf{mul} instruction, adding the result to an accumulator
with \textbf{add}, and handling loop overhead and memory loads.
By introducing a CXU instruction that combines multiplication and accumulation into a
single hardware step, the instruction footprint and pipeline hazards within the inner
inference loop can be substantially reduced.
The MAC extension and its evaluation within TFLite Micro are described in
Sections~\ref{subsec:tflite_impl} and~\ref{subsec:tflite_results}.
